---
title: "Code for 73274 Econometrics 1 Final Project"
author: "Gordon Wang"
date: "May 11, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r}
# Read in data
library(foreign)
bw.data = read.dta("birthweight_smoking.dta")

# Check for NAs
sum(is.na(bw.data))

# Data summary
dim(bw.data)
summary(bw.data)
```


# Univariate EDA
```{r}
# 1-Dimensional EDA on Birthweight and Smoking, the main variables of interest
par(mfrow = c(2, 2))
hist(bw.data$birthweight, main = "Distribution of Infant's Birthweight", xlab = "Birthweight of Infant (g)", ylab = "Frequency", col = "red")

smoker_categories = table(as.factor(bw.data$smoker))
names(smoker_categories) = c("Didn't Smoke", "Smoked")
smoker_categories
barplot(smoker_categories, main = "Barplot of Mother Smoking Variable", xlab = "Smoked During Pregnancy", ylab = "Frequency")


# 1-Dimensional EDA on Mother's Attributes
hist(bw.data$age, main = "Distribution of Mother's Age", xlab = "Age", ylab = "Frequency", col = "blue")

hist(bw.data$educ, main = "Distribution of Mother's Education", xlab = "Education (years)", ylab = "Frequency", col = "cornflowerblue")

unmarried_categories = table(as.factor(bw.data$unmarried))
names(unmarried_categories) = c("Married", "Unmarried")
unmarried_categories
barplot(unmarried_categories, main = "Barplot of Mother Unmarried", xlab = "Mother's Marriage Status", ylab = "Frequency")


# 1-Dimensional EDA on This Pregnancy
alcohol_categories = table(as.factor(bw.data$alcohol))
names(alcohol_categories) = c("No Alcohol", "Alcohol")
alcohol_categories
barplot(alcohol_categories, main = "Barplot of Alcohol", xlab = "Drank Alcohol During Pregnancy", ylab = "Frequency")


hist(bw.data$drinks, main = "Distribution of Drinks", xlab = "Number of Drinks/Week", ylab = "Frequency", col = "orange")
drinks_categories = table(as.factor(bw.data$drinks))
drinks_categories
barplot(drinks_categories, main = "Barplot of Number of Drinks/Week", xlab = "Drinks/Week", ylab = "Frequency")

tripre1_categories = table(as.factor(bw.data$tripre1))
names(tripre1_categories) = c("No", "Yes")
tripre1_categories
barplot(tripre1_categories, main = "Barplot of Tripre1", xlab = "1st Prenatal Care Visit in 1st Trimester", ylab = "Frequency")

tripre2_categories = table(as.factor(bw.data$tripre2))
names(tripre2_categories) =  c("No", "Yes")
tripre2_categories
barplot(tripre2_categories, main = "Barplot of Tripre2", xlab = "1st Prenatal Care Visit in 2nd Trimester", ylab = "Frequency")

tripre3_categories = table(as.factor(bw.data$tripre3))
names(tripre3_categories) =  c("No", "Yes")
tripre3_categories
barplot(tripre3_categories, main = "Barplot of Tripre3", xlab = "1st Prenatal Care Visit in 3rd Trimester", ylab = "Frequency")

tripre0_categories = table(as.factor(bw.data$tripre0))
names(tripre0_categories) =  c("No", "Yes")
tripre0_categories
barplot(tripre0_categories, main = "Barplot of Tripre0", xlab = "No Prenatal Visits", ylab = "Frequency")

hist(bw.data$nprevist, main = "Distribution of Prenatal Visits", xlab = "Total Number of Prenatal Visits", ylab = "Frequency", col = "blue")
```

# Bivariate EDA
```{r}
# Boxplot for variables of interest
boxplot(bw.data$birthweight ~ bw.data$smoker, names = c("Didn't Smoke", "Smoked"), main = "Birthweight vs Smoking", xlab = "Mother Smoking During Pregnancy", ylab = "Infant Birthweight (g)", col = "slategray4")

# Pairs plot for continuous variables
pairs(bw.data[,c(7,10,11,1,12)], pch = ".")
```


# Regressions

# Regressing Y on main x of interest
```{r}
# Regressing Birthweight on Smoker
model1 = lm(birthweight ~ smoker, data = bw.data)
summary(model1)
```

# Controlling for All other variables
```{r}
# Start with all regressors in the model
model.full = lm(birthweight ~ ., data = bw.data)
summary(model.full)

# Drop tripre0 from model to avoid dummy variable trap
model2 = lm(birthweight ~ .-tripre0, data = bw.data)
summary(model2)
```

Tripre0 has a NA slope coefficient. This is because of the dummy variable trap caused by perfect collinearity among tripre1, tripre2, tripre3, and tripre0. The number of 0 values for tripre0 equals the sum of 1 values for tripre1, tripre2, and tripre3. Perfect collinearity is a violation of Multiple Linear Regression Assumption 3. Thus, the decision was made to drop tripre0 from the model.


```{r}
# Eliminate redundacy from the model by dropping alcohol or drinks
model3 = lm(birthweight ~.-tripre0-alcohol,data= bw.data)
summary(model3)

model3b = lm(birthweight ~. -tripre0-drinks, data = bw.data)
summary(model3b)
```

I had to drop either the drinks or alcohol variable because of redundancy. The number of mothers who didn't drink once per week is equal to the number of mothers who didn't drink alcohol, coded as 0. To decide which variable to drop, I compared the multiple R-squared values of the model upon excluding one or the other, and found out that dropping alcohol resulted in a higher value than dropping drinks. Thus, I decided to drop the alcohol variable from the model.

```{r}
# Check variance inflation factors of model
library(car)
vif(model3)

# Fit model without highly correlated predicters
model.reduced = lm(birthweight ~.-tripre0-alcohol-tripre1-tripre2-tripre3,data = bw.data)
summary(model.reduced)

# Test for joint significance of tripre1, tripre2, and tripre3
anova(model.reduced, model3, test = "F")
```

```{r}
# Try dropping insignificant predictors from model
model4 = lm(birthweight ~ smoker + unmarried + nprevist + tripre1 + tripre2 + tripre3, data = bw.data)
summary(model4)

# Compare Model4 to previous model using F-test
anova(model4, model3, test = "F")
```

```{r}
# Test for interactions
model.unrestricted = lm(birthweight ~ smoker + unmarried + nprevist + tripre1 + tripre2 + tripre3 + smoker*nprevist + smoker*unmarried + smoker*tripre1 + smoker*tripre2 + smoker*tripre3, data = bw.data)
summary(model.unrestricted)

# F-Test for whether smoker's interactions are jointly significant
anova(model4, model.unrestricted, test = "F")

# Fit a model only including the significant interaction
model5 = lm(birthweight ~ smoker + unmarried + nprevist + tripre1 + tripre2 + tripre3 + smoker*unmarried, data = bw.data)
summary(model5)

# F-test for excluding all individually insignificant interactions
anova(model5, model.unrestricted, test = "F")
```

```{r}
# Selected model, without accounting for diagnostics
final.mod = lm(birthweight ~ smoker + unmarried + nprevist + tripre1 + tripre2 + tripre3 + smoker*unmarried, data = bw.data)
summary(final.mod)

# Diagnostics for Chosen Model
plot(fitted(final.mod), resid(final.mod), pch = ".", main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residual Values")
abline(h = 0, lty = 'dotdash', col = 'red')

hist(final.mod$residuals)
qqnorm(residuals(final.mod), pch= ".")
qqline(residuals(final.mod))


plot(final.mod, pch = ".")
```


```{r}
# Run a heteroskedasticity test
library(car)
ncvTest(final.mod)
```
P-value less than 0.05, so we reject the null hypothesis of that the variance of the residuals is constant and infer that heteroskedasticity is present.


